---
title: "Smoothing Model Building"
author: "Donald Duck"
output: 
    pdf_document
---

* UW ID: 20654321
* Kaggle public score: ???
* Kaggle private score: ???
<!-- The number of submissions you made to Kaggle -->
* Kaggle submission count/times: ??

Important: follow the structure of the document and summarize important handlings in the summary section. The structure here is kept at a minimum, feel free if you have more subsections and details to add.

# Summary

## Preprocessing

### Transformation (if any, delete if none)
* price: I squared the response variate price.

### New Variables (if any, delete if none)
<!-- List all variables/predictors added to dtrain and dtest  -->
* isGood: indicator of whether the condition is "Good".


### Missing data handling
<!-- There are missing values in some variables, need to do something -->
* yr_rmdl: replace the missing values in yr_rmdl with median.

## Model Building

Main package used: `magicSpline`

Forward selection and adding terms using thin plate spline...

## Final Model
<!-- formula in your final fitted function  -->
* The final model is $price \sim$ fireplaces + isGood

<!-- Details -->
<!-- R code starts, please present both code and output. -->
<!-- Provide descriptions and explanations to justify your operations -->
<!-- Be brief and to the point. -->
<!-- Only details lead to your final model are needed. If you tried other appoaches/models, you can mention them, no details needed for abandoned attempts. -->



# 1.Preprocessing

## 1.1 Loading data
```{r,echo=TRUE}
load("smooth.Rdata")
```

## 1.2 Missing data handling
```{r,echo=TRUE}
```

# 2. Model building
```{r,echo=TRUE}
# to add a term, or not, that is the question
```

